\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-09-29 paper.qmd
Added the Figure 5 title

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data ($p\text{-}D$) by applying a non-linear transformation. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of $p\text{-}D$ distributions. The NLDR methods and (hyper)parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the $p\text{-}D$ data, we have developed an algorithm to show the \gD{} NLDR model in the $p\text{-}D$ space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

@@

Non-linear dimension reduction (NLDR) is popular for making a convenient low-dimensional (\kD{}) representation of high-dimensional (\pD{}) data. Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2022], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021]. However, the representation generated can vary dramatically from method to method, and with different choices of parameters or random seeds made using the same method (@fig-NLDR-variety). The dilemma for the analyst is then, **which representation to use**. The choice might result in different procedures used in the downstream analysis, or different inferential conclusions. The research described here provides new visual tools to aid with this decision.

@@

The paper is organised as follows. @sec-background provides a summary of the literature on NLDR, and high-dimensional data visualization methods. @sec-method contains the details of the new methodology, including simulated data examples. Two applications illustrating the use of the new methodology for bioinformatics and image classification are in @sec-applications. Limitations and future directions are provided in @sec-discussion.

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-umap1
#| fig-cap: "Model in \\gD{}, on the layout and two views of the fit in projections from $7\\text{-}D$, for the two-curvy clusters data ($n =  2000$ and $p = 7$). The fitted model"
#| fig-width: 10
#| fig-height: 10

>>>>

---
title: "Looking at Non-Linear Dimension Reductions as Models in the Data Space"
format: 
    jasa-pdf:
        keep-tex: true
    jasa-html: default
author:
  - name: Jayani P.G. Lakshika
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-6265-6481
    email: jayani.piyadigamage@monash.edu
    url: https://jayanilakshika.netlify.app/
  - name: Dianne Cook
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3813-7155
    email: dicook@monash.edu 
    url: http://www.dicook.org/
  - name: Paul Harrison
    affiliations:
      - name: Monash University
        department: MGBP, BDInstitute
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0002-3980-268X
    email: 	paul.harrison@monash.edu
    url: 
  - name: Michael Lydeamore
    affiliations:
      - name: Monash University
        department: Econometrics & Business Statistics
        address: Clayton
        city: VIC 
        country: Australia
        postal-code: 3800
    orcid: 0000-0001-6515-827X
    email: michael.lydeamore@monash.edu
    url: 
  - name: Thiyanga S. Talagala
    affiliations:
      - name: University of Sri Jayewardenepura
        department: Statistics
        address: Gangodawila
        city: Nugegoda 
        country: Sri Lanka
        postal-code: 10100
    orcid: 0000-0002-0656-9789
    email: ttalagala@sjp.ac.lk 
    url: https://thiyanga.netlify.app/
tbl-cap-location: bottom
abstract: |
  Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP provide a low-dimensional representation of high-dimensional data (\pD{}) by applying a non-linear transformation. NLDR often exaggerates random patterns, sometimes due to the samples observed. But NLDR views have an important role in data analysis because, if done well, they provide a concise visual (and conceptual) summary of \pD{} distributions. The NLDR methods and (hyper)parameter choices can create wildly different representations, making it difficult to decide which is best, or whether any or all are accurate or misleading. To help assess the NLDR and decide on which, if any, is the most reasonable representation of the structure(s) present in the \pD{} data, we have developed an algorithm to show the \gD{} NLDR model in the \pD{} space, viewed with a tour, a movie of linear projections. From this, one can see if the model fits everywhere, or better in some subspaces, or completely mismatches the data. Also, we can see how different methods may have similar summaries or quirks.

@@

Non-linear dimension reduction (NLDR) is popular for making a convenient low-dimensional (\kD{}) representation of high-dimensional (\pD{}) data ($k < p$). Recently developed methods include t-distributed stochastic neighbor embedding (tSNE) [@laurens2008], uniform manifold approximation and projection (UMAP) [@leland2018], potential of heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm [@moon2019], large-scale dimensionality reduction Using triplets (TriMAP) [@amid2022], and pairwise controlled manifold approximation (PaCMAP) [@yingfan2021]. However, the representation generated can vary dramatically from method to method, and with different choices of parameters or random seeds made using the same method (@fig-NLDR-variety). The dilemma for the analyst is then, **which representation to use**. The choice might result in different procedures used in the downstream analysis, or different inferential conclusions. The research described here provides new visual tools to aid with this decision.

@@

The paper is organized as follows. @sec-background provides a summary of the literature on NLDR, and high-dimensional data visualization methods. @sec-method contains the details of the new methodology, including simulated data examples. Two applications illustrating the use of the new methodology for bioinformatics and image classification are in @sec-applications. Limitations and future directions are provided in @sec-discussion.

@@

<!-- langevitour with pD model for two-curvy clusters -->
```{r}
#| label: best-fit-umap1
#| fig-cap: "Model in \\gD{} ($a_1 = 0.12$), on the layout and two views of the fit in projections from $7\\text{-}D$, for the two-curvy clusters data ($n =  2000$ and $p = 7$). One view shows the fitted model, and the other shows the fitted and true models alongside the data. The fitted model accurately represents the shape of the clusters but does not capture the edges of the structure. This illustrates a characteristic of UMAP, which compresses the data when transforming the data into \\gD{}."
#| fig-width: 10
#| fig-height: 10

### 2024-09-29 paper.tex
Added the Figure 5 title

\bigskip
\bigskip
\begin{abstract}
Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP
provide a low-dimensional representation of high-dimensional data
(\(p\text{-}D\)) by applying a non-linear transformation. NLDR often
exaggerates random patterns, sometimes due to the samples observed. But
NLDR views have an important role in data analysis because, if done
well, they provide a concise visual (and conceptual) summary of
\(p\text{-}D\) distributions. The NLDR methods and (hyper)parameter
choices can create wildly different representations, making it difficult
to decide which is best, or whether any or all are accurate or
misleading. To help assess the NLDR and decide on which, if any, is the
most reasonable representation of the structure(s) present in the
\(p\text{-}D\) data, we have developed an algorithm to show the \gD{}
NLDR model in the \(p\text{-}D\) space, viewed with a tour, a movie of
linear projections. From this, one can see if the model fits everywhere,
or better in some subspaces, or completely mismatches the data. Also, we
can see how different methods may have similar summaries or quirks.
\end{abstract}

@@

Non-linear dimension reduction (NLDR) is popular for making a convenient
low-dimensional (\kD{}) representation of high-dimensional (\pD{}) data.
Recently developed methods include t-distributed stochastic neighbor
embedding (tSNE) \citep{laurens2008}, uniform manifold approximation and
projection (UMAP) \citep{leland2018}, potential of heat-diffusion for
affinity-based trajectory embedding (PHATE) algorithm \citep{moon2019},
large-scale dimensionality reduction Using triplets (TriMAP)
\citep{amid2022}, and pairwise controlled manifold approximation
(PaCMAP) \citep{yingfan2021}. However, the representation generated can
vary dramatically from method to method, and with different choices of
parameters or random seeds made using the same method
(Figure~\ref{fig-NLDR-variety}). The dilemma for the analyst is then,
\textbf{which representation to use}. The choice might result in
different procedures used in the downstream analysis, or different
inferential conclusions. The research described here provides new visual
tools to aid with this decision.

@@

The paper is organised as follows. Section~\ref{sec-background} provides
a summary of the literature on NLDR, and high-dimensional data
visualization methods. Section~\ref{sec-method} contains the details of
the new methodology, including simulated data examples. Two applications
illustrating the use of the new methodology for bioinformatics and image
classification are in Section~\ref{sec-applications}. Limitations and
future directions are provided in Section~\ref{sec-discussion}.

@@

\caption{Model in \gD{}, on the layout and two views of the fit in
projections from \(7\text{-}D\), for the two-curvy clusters data
(\(n =  2000\) and \(p = 7\)). The fitted model}

>>>>

\bigskip
\bigskip
\begin{abstract}
Non-linear dimension reduction (NLDR) techniques such as tSNE, and UMAP
provide a low-dimensional representation of high-dimensional data
(\pD{}) by applying a non-linear transformation. NLDR often exaggerates
random patterns, sometimes due to the samples observed. But NLDR views
have an important role in data analysis because, if done well, they
provide a concise visual (and conceptual) summary of \pD{}
distributions. The NLDR methods and (hyper)parameter choices can create
wildly different representations, making it difficult to decide which is
best, or whether any or all are accurate or misleading. To help assess
the NLDR and decide on which, if any, is the most reasonable
representation of the structure(s) present in the \pD{} data, we have
developed an algorithm to show the \gD{} NLDR model in the \pD{} space,
viewed with a tour, a movie of linear projections. From this, one can
see if the model fits everywhere, or better in some subspaces, or
completely mismatches the data. Also, we can see how different methods
may have similar summaries or quirks.
\end{abstract}

@@

Non-linear dimension reduction (NLDR) is popular for making a convenient
low-dimensional (\kD{}) representation of high-dimensional (\pD{}) data
(\(k < p\)). Recently developed methods include t-distributed stochastic
neighbor embedding (tSNE) \citep{laurens2008}, uniform manifold
approximation and projection (UMAP) \citep{leland2018}, potential of
heat-diffusion for affinity-based trajectory embedding (PHATE) algorithm
\citep{moon2019}, large-scale dimensionality reduction Using triplets
(TriMAP) \citep{amid2022}, and pairwise controlled manifold
approximation (PaCMAP) \citep{yingfan2021}. However, the representation
generated can vary dramatically from method to method, and with
different choices of parameters or random seeds made using the same
method (Figure~\ref{fig-NLDR-variety}). The dilemma for the analyst is
then, \textbf{which representation to use}. The choice might result in
different procedures used in the downstream analysis, or different
inferential conclusions. The research described here provides new visual
tools to aid with this decision.

@@

The paper is organized as follows. Section~\ref{sec-background} provides
a summary of the literature on NLDR, and high-dimensional data
visualization methods. Section~\ref{sec-method} contains the details of
the new methodology, including simulated data examples. Two applications
illustrating the use of the new methodology for bioinformatics and image
classification are in Section~\ref{sec-applications}. Limitations and
future directions are provided in Section~\ref{sec-discussion}.

@@

\caption{Model in \gD{} (\(a_1 = 0.12\)), on the layout and two views of
the fit in projections from \(7\text{-}D\), for the two-curvy clusters
data (\(n =  2000\) and \(p = 7\)). One view shows the fitted model, and
the other shows the fitted and true models alongside the data. The
fitted model accurately represents the shape of the clusters but does
not capture the edges of the structure. This illustrates a
characteristic of UMAP, which compresses the data when transforming the
data into \gD{}.}
